---
title: "HW Class12 Pt.2 (Population analysis) [Q13 Q14 BoxPlot]"
author: "Nhi To (A18053310)"
date: 2/17/2024
tcoc: TRUE
format: pdf
---

## Section 1. Proproportion og G/G in a population 

Download a CSV file from Ensemble < https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39830097-39960098;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel

Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.)/ nrow(mxl)
```

Now let's look at a different population. I picked the GBR

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")

```

Find a proportion of G G

```{r}
table(gbr$Genotype..forward.strand.)/nrow(gbr)
```

This variant that is associated with childhood asthma is more frequent in the GBR population with the MKL population. 

Let's now dig into this further


## Section 4: Population Scale Analysis

One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3
expression.

>Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.  Hint: The read.table(), summary() and boxplot() functions will likely be useful here. There is an example R script online to be used ONLY if you are struggling in vein. Note that you can find the medium value from saving the output of the boxplot() function to an R object and examining this object. There is also the medium() and summary() function that you can use to check your understanding.  

>**ANSWER:** The sample size for A/A genotype is 108. The sample size for A/G genotype is 233. The sample size for G/G genotype is 121. The corresponding median expression level for A/A genotype is 31.24847, A/G is 25.06486, and G/G is 20.07363.


```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```


```{r}
table(expr$geno)
```

```{r}
genotype_median <- aggregate(exp ~ geno, data = expr, FUN = median)
genotype_median
```


```{r}
summary(expr)
```



>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3? 

>**ANSWER**: The boxplot with a box per genotype is generated below. From the boxplot, we can visually see that the A/A genotype has the highest median expression of ORMDL3, while G/G has the lowest. Since the expression of ORMDL3 is lowest when the genotype is G/G, we can infer that the G allele is associated with the decrease of ORMDL3 expression. Yes, the SNP (single nucleotide polymorphism), has an effect on the expression of ORMDL3. Having an A allele will have some type of increasing effect on the expression of ORMDL3, while the G allele will have a decreasing effect. This suggests that SNP regulate the expression of ORMDL3.  


```{r}
library(ggplot2)
```


Let's make a boxplot

```{r}
ggplot(expr) +
  aes(geno, exp, fill=geno) +
  geom_boxplot(notch=TRUE)
```










