---
title: "Halloween Candy Mini Project"
author: "Nhi To"
format: pdf
date: February 4, 2025
toc: true
---


Today we will examine data from 538 on common Halloween candy. In particular we will use ggplot, dplyr, and PCA to make sense of this multivariate dataset

## Importing candy data

```{r}
candy <- read.csv("candy-data copy.csv", row.names=1)
head(candy)
```

>Q1. How many different candy types are in this dataset?

>**ANSWER**: There are 85 different candy types in this dataset.

```{r}
nrow(candy)
```

>Q2. How many fruity candy types are in the dataset?

>**ANSWER**: There are 38 fruity candy types in the dataset.

```{r}
sum(candy$fruity)

```

>Additional Question: How many chcocolate candy are there in the dataset?

>**ANSWER**: There are 37 chocolate candy in the dataset.

```{r}
sum(candy$chocolate)
```


>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

>**ANSWER**: My favorite candy is Pop Rocks, and its winpercent is 41.26551.

```{r}
candy["Pop Rocks", "winpercent"]
```


>Q4. What is the winpercent value for “Kit Kat”?

>**ANSWER**: The winpercent for Kit Kat is 76.7686.

```{r}
candy["Kit Kat", "winpercent"]
```

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

>**ANSWER**: The winpercent value for Tootsie Roll Snack Bars is 49.6535.

```{r}
candy["Tootsie Roll Snack Bars", "winpercent"]
```

```{r}
library("skimr")
skim(candy)
```


>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

>**ANSWER**: The variable 'winpercent' appears to be on a different scale compared to the majority of the other columns in the dataset, because the other variables are between 0-1. The variable 'winpercent' values are between 0-100%, rather than 0-1, revealing that it is at a different scale value. Therefore, I will need to scale this dataset before analysis like PCA. 

>Q7. What do you think a zero and one represent for the candy$chocolate column?

>**ANSWER**: I think that the zero represents that that particular candy does not have any chocolate inside it. A one would represent that that candy contains chocolate. 

>Q8. Plot a histogram of winpercent values

>**ANSWER**:

```{r}
library(ggplot2)

ggplot(candy)+
  aes(winpercent)+
  geom_histogram(bins=10, fill="lightpink")
```


```{r}
hist(candy$winpercent)
```

>Q9. Is the distribution of winpercent values symmetrical?

>**ANSWER**:The distribution of winperent values are  slightly right skewed, and therefore not symmetrical. 

>Q10. Is the center of the distribution above or below 50%?

>**ANSWER**: The center of the distribution is below 50%. Visually, based on the peak of the data, the center is around 40%. Based on the summary below, the median value is 47.83, which is below 50%.

```{r}
summary(candy$winpercent)
```


>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

-Step 1: find all "chocolate" candy
-Step 2: find their "winpercent" values
-Step 3: Summarize these values
-Step 4: find all "fruity" candy
-Step 5:find their winpercent values
-Step 6: Summarize these values
-Step 7: Compare the two summary values.

>**ANSWER**: The mean value for chocolate candy is 60.92153 compared to the mean value of the fruity candy at 44.11974. Therefore we conclude that, on average, chocolate candy higher than fruity candy.

Step 1
```{r}
choc.inds <- candy$chocolate == 1
```

Step 2
```{r}
choc.win <- candy[choc.inds,]$winpercent
```

Step 3
```{r}
mean(choc.win)
```

Step 4
```{r}
fruit.inds <- candy$fruity == 1
```

Step 5:
```{r}
fruit.win <- candy[fruit.inds,]$winpercent
```

Step 6:
```{r}
mean(fruit.win)
```

Step 7:
```{r}
mean(choc.win) > mean(fruit.win)
```

>Q12. Is this difference statistically significant?

>**ANSWER**:The p-value (2.871e-08) is less than 0.05, indicating that this difference is statistically signficant.

```{r}
t.test(choc.win, fruit.win)
```

>Q13. What are the five least liked candy types in this set?

>**ANSWER**: The five least liked candy types in this set is the Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters.

```{r}
sort(candy$winpercent)
```


>Q14. What are the top 5 all time favorite candy types out of this set?

>**ANSWER**: The top 5 all time favorite candy types out of this set is Reese's Peanut Butter cup, Reese's miniatures, Twix, Kit Kat, and Snickers. 

```{r}
# Not that useful- it just sorts the values
sort(candy$winpercent)
```

```{r}
x<- c(10, 1, 100)
sort(x)
order(x)
```

```{r}
x[order (x)]
```

The 'order()' function tells us how to arrange the elments of the input to make them sorted - i.e. how to order them

We can determine the order of winpercent to make sorted and use them sorted and use that order to arrange the whole dataset.

```{r}
ord.inds <- order(candy$winpercent)
head( candy[ord.inds,] )
```

```{r}
ord.inds <- order(candy$winpercent, decreasing=T)
head( candy[ord.inds,] )
```


>Q15. Make a first barplot of candy ranking based on winpercent values.

>**ANSWER**:

```{r}
ggplot(candy)+
  aes(winpercent, rownames(candy))+
  geom_col()
```


>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

>**ANSWER**:

```{r}
ggplot(candy)+
  aes(winpercent, reorder(rownames(candy), winpercent), fill=chocolate)+
  geom_col() 
```

We need to make our own separate color vector where we can spell out what candy is colored a particular color

```{r}
mycols <- rep("gray", nrow(candy))
mycols[candy$chocolate == 1] <- "chocolate"
mycols[candy$bar == 1] <- "red"
mycols[candy$fruity == 1] <- "blue"
mycols
```

```{r}
ggplot(candy)+
  aes(winpercent, reorder(rownames(candy), winpercent))+
  geom_col(fill=mycols) 
```

Now, for the first time, using this plot we can answer questions like:
>Q17. What is the worst ranked chocolate candy?

>**ANSWER**: The worst ranked chcolate candy is Sixlets.

>Q18. What is the best ranked fruity candy?

>**ANSWER**: The best ranked fruity candy is Starburst.

## Taking a look at pricepercent

Make a plot of winpercent (x-axis) vs pricepercent (y-axis)

```{r}
ggplot(candy)+
  aes(winpercent, pricepercent, label=rownames(candy))+ 
  geom_point(col=mycols)+
  geom_text()
```

To avoid the overplotting of the text label, we can use the add on package **ggrepel**

```{r}

library (ggrepel)

ggplot(candy)+
  aes(winpercent, pricepercent, label=rownames(candy))+ 
  geom_point(col=mycols)+
  geom_text_repel(max.overlaps= 6)+
  theme_bw()
```

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

>**ANSWER**: The candy type that is the highest ranked in terms of winpercent for the least money is Tootsie Roll Midgies. The highest winpercent/priceperent ratio is 4157.8862, which is the for the Tootsie Roll Midgies.


```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )

candy$Q19 <- candy$winpercent/candy$pricepercent

ord2<- order(candy$Q19, decreasing=TRUE)
head( candy[ord2, ], n=5 )
```

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

>**ANSWER**: The top 5 most expensive candy types in the dataseet are Nik L Nip, Nestle Smarties, Ring pop, Hershe's Krackel, and Hershey's Milk Chocolate. Out of these 5, Nik L Nip has the lowest winpercent, and thus is the least popular out of the 5 most expensive candy types. 


## 5 Exploring the correlation structure

Now that we have explored the dataset a little, we will see how the variables interact with one another.

First, we wil use correlation and view the results with the **corplot** package to plot a correlation matrix

```{r}
cij <- cor(candy)
```

```{r}
library(corrplot)

corrplot(cij)
```


>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

>**ANSWER**: Based on this plot, the two variables "fruity" and "chocolate" are anti-correlated, because they both have negative values (depicted by the red circle). This means that fruity candy tends to not have chocolate in it.

>Q23. Similarly, what two variables are most positively correlated?

>**ANSWER**: Based on this plot, the two variables "chocolate" and "caramel" are positively correlated, because they both have positive vlaues (depicted by the blue circle). This means that chocolate candy tends to have caramel in it.

## 6. Principal Component Analysis 

>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

>**ANSWER**: The original variables that are picked up by PC1 in the positive direction are the fruity, pluribis, and hard. This means that that fruity candy are often pluribis and hard as well. 

Let's apply PCA using the 'prcomp()' function


```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
attributes(pca)
```

Let's plot our main results as our PCA "score plot"

```{r}

ggplot(pca$x)+
  aes(PC1, PC2, label=rownames(pca$x))+
  geom_point(col=mycols)+
  geom_text_repel(col=mycols)

```

Finally let's look at how the original variables contribute to the PCs, start with PC1

```{r}
pca$rotation
```

```{r}
ggplot(pca$rotation)+
  aes(PC1, reorder(rownames(pca$rotation),PC1))+ 
  geom_col()
```




